Inference using context-dependent models of mutation

### 2021-10-11 writeups/context-dependence.tex
The nit-picking will not stop until morale improves

  \noindent
  This only has one parameter.
  If we allow particles to exit from the right, and new particles to enter from the left
  then from observing only starting and ending configurations
  we cannot tell with certainty which particles have moved where,
  so it is not obvious how to estimate the speed.
  However, summing over possible particle movements would allow us to
  compute the likelihood of a given configuration change,
  and so estimate the speed by maximum likelihood.

@@

As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
for each position $i$, let $J(i,x,y)$ be those transitions in $\calT$ that can be applied at $i$
to change $x$ into $y$, i.e.,
\[
%EM I think we want h and not j here in the parenthetical superscript.
%EM Perhaps we want \mu^{j} rather than \mu^{(j)}? I think the latter is more for subsequences.
%EM And do we want to define that j is some meta-index for all triples?
    J(i,x,y) = \{ (\mu^{(j)},u^{(j)},v^{(j)}) \in \calT \st x \xrightarrow{i,u^{(j)},v^{(j)}} y \}.
\]
The rate ${G(n)}_{x,y}$ is then the sum of all matching transition rates,
multiplied by the fitness term,
namely
\begin{align} \label{eqn:G_defn}
%EM Are we happy with leaving the \sum_{J(i,x,y)} a little vague here? I think it's "j" for which there is a triple?
    {G(n)}_{x,y} = \phi\left(E(y)-E(x)\right) \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then ${G(n)}_{x,y}=0$.
%EM Do we need this following sentence? If so, we need to update it to also say something about the fitness term.
This formalizes the statement that the transition rate from $x$ to $y$ is the sum of the rates of all transition triples
that would change $x$ into $y$.
In principle, this gives us the transition probabilities for the process by a matrix exponential:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = {\left(e^{tG(n)}\right)}_{xy} ,
\end{align}
which would then provide a route to parameter estimation.
%EM Here or somewhere else do we need to say something about setting the overall rate of the process? In phylogenetics it is common to scale the rate matrix so that substitutions happen with rate 1.
In practice, the size of these matrices makes direct application obviously impractical for anything but very small $n$.
This paper proposes an alternate strategy.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''
or ``uniformization'' \citep{hobolth2009simulation}.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu^*$ per site,
and then resolving each possible change in temporal order.
The rate $\mu^*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
    \mu^* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu^{(j)} \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
Now sample possible changes:
in a sequence of length $L$, over a time period $t$,
first draw the number of possible changes, $N$,
from a Poisson distribution with mean $\mu^* t L$,
and then choose the times $s_k$ and locations $i_k$
at which possible changes occur uniformly
over $[0, t]$ and $\{1, 2, \ldots, L\}$ respectively.
Order the possible change times, so that
\[
    0 < s_1 < \cdots < s_N < t .
\]
(Note that this is equivalent to saying that possible
changes occur at rate $\mu^*$ independently at each of the $L$ sites.)
Now, suppose we have determined the state just before time $s_k$ to be $X(s_k) = x$.
Then, a new state is then chosen at position $i_k$
with probability proportional to the corresponding rate, i.e.,
out of the possible transitions $\calT$, the $j^\text{th}$ one is chosen
with probability
\[
q_j = \begin{cases}
    \mu^{(j)}/\mu^* \qquad & \text{if } x_i^{(|u^{(j)}|)} = u^{(j)}  \\
    0 \qquad & \text{otherwise.}
\end{cases}
\]
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$
gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0\ge 0$.)
If triple $(\mu^{(j)}, u^{(j)}, v^{(j)})$ is chosen, then
the substring of $x$ matching $u^{(j)}$ that begins at position $i$
is replaced with $v^j$ to produce $X(s_k)$.

>>>>

  \noindent
  This only has one parameter.
  If we allow particles to exit from the right, and new particles to enter from the left
  then from observing only starting and ending configurations
  we cannot tell with certainty which particles have moved where,
  so it is not obvious how to estimate the speed.
  However, summing over possible particle movements would allow us to
  compute the likelihood of a given configuration change,
  and so estimate the speed by maximum likelihood
  given an initial sequence, a final sequence and an elapsed time.

@@

As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
for each position $i$, let $J(i,x,y)$ be those transitions in $\calT$ that can be applied at $i$
to change $x$ into $y$, i.e.,
\[
%EM I think we want h and not j here in the parenthetical superscript.
%EM Perhaps we want \mu^{j} rather than \mu^{(j)}? I think it could be clearer to keep the latter just for subsequences. How about lower-j?
%EM And do we want to define that j is some meta-index for all triples?
    J(i,x,y) = \{ (\mu^{(j)},u^{(j)},v^{(j)}) \in \calT \st x \xrightarrow{i,u^{(j)},v^{(j)}} y \}.
\]
The rate ${G(n)}_{x,y}$ is then the sum of all matching transition rates,
multiplied by the fitness term,
namely
\begin{align} \label{eqn:G_defn}
%EM Are we happy with leaving the \sum_{J(i,x,y)} a little vague here? I think it's "j" for which there is a triple?
    {G(n)}_{x,y} = \phi\left(E(y)-E(x)\right) \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then ${G(n)}_{x,y}=0$.
%EM Do we need this following sentence? If so, we need to update it to also say something about the fitness term.
This formalizes the statement that the transition rate from $x$ to $y$ is the sum of the rates of all transition triples
that would change $x$ into $y$.
In principle, this gives us the transition probabilities for the process by a matrix exponential:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = {\left(e^{tG(n)}\right)}_{xy} ,
\end{align}
which would then provide a route to parameter estimation.
%EM Here or somewhere else do we need to say something about setting the overall rate of the process? In phylogenetics it is common to scale the rate matrix so that substitutions happen with rate 1.
In practice, the size of these matrices makes direct application obviously impractical for anything but very small $n$.
This paper proposes an alternate strategy.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''
or ``uniformization'' \citep{hobolth2009simulation}.
%EM I note that here we are re-using the "possible change" terminology, which we used above for something that seems slightly different, namely the fitness setup.
%EM It seems like either we should fit these possible changes into that framework, or call these something else.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu^*$ per site,
and then resolving each possible change in temporal order.
The rate $\mu^*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
    %EM If we change anything about j and J above, propagate that change here.
    \mu^* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu^{(j)} \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
Now sample possible changes:
%EM Do we want to have sequence length become L here? It has been $n$ so far.
in a sequence of length $L$, over a time period $t$,
first draw the number of possible changes, $N$,
from a Poisson distribution with mean $\mu^* t L$,
and then choose the times $s_k$ and locations $i_k$
at which possible changes occur uniformly
over $[0, t]$ and $\{1, 2, \ldots, L\}$ respectively.
Order the possible change times, so that
\[
    0 < s_1 < \cdots < s_N < t .
\]
(Note that this is equivalent to saying that possible
changes occur at rate $\mu^*$ independently at each of the $L$ sites.)
Now, suppose we have determined the state just before time $s_k$ to be $X(s_k) = x$.
Then, a new state is then chosen at position $i_k$
with probability proportional to the corresponding rate, i.e.,
out of the possible transitions $\calT$, the $j^\text{th}$ one is chosen
with probability
\[
q_j = \begin{cases}
    %EM Perhaps we could look at this notation together? I would think that u^{(j)} is the length-j subsequence of u, but I don't think that's what's desired here.
    \mu^{(j)}/\mu^* \qquad & \text{if } x_i^{(|u^{(j)}|)} = u^{(j)}  \\
    0 \qquad & \text{otherwise.}
\end{cases}
\]
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$
gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0\ge 0$.)
If triple $(\mu^{(j)}, u^{(j)}, v^{(j)})$ is chosen, then
the substring of $x$ matching $u^{(j)}$ that begins at position $i$
is replaced with $v^j$ to produce $X(s_k)$.
